---
title: "Class 17 vaccination"
author: "dairabel"
format: pdf
---
# Background

For todays class we will explore the data set on state wide vaccination rates from CA.Gov

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

## Data entry

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)

```
> Q1. What column details the total number of people fully vaccinated?

```{r}
head(vax$persons_fully_vaccinated)
```
A: The column for total number of people fully vaccinated is the persons_fully_vaccinated column.

> Q2. What column details the Zip code tabulation area?

```{r}
head(vax$zip_code_tabulation_area)
```
A: The second column, the zip_code_tabulation_area

>Q3. What is the earliest date in this dataset?

```{r}
head(vax$as_of_date[1])
```
A: The earliest was on January 5th, 2021.

>Q4. What is the latest date in this dataset?

```{r}
tail(vax$as_of_date)
```
A: Last one is March 07, 2023.


Useful function for exploring the data! use skimr package **skimr**

```{r}
library(skimr)
skim(vax)
#or any package
#skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

A: there are 13 numeric, based on skmir results

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
n.missing
```
A: There are 16621 NA values in the persons_fully_vaccinated  column

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round(n.missing/nrow(vax) * 100, 2)
```
A: 8.27 % are missing.

> Q8[Optional]: Why might this data be missing?

A: One reason is that it was at the early stages of vaccination efforts and the tracking wasn't standard or it was new.

## Working with the dates

We will use the **lubridate** package to help ease the pain of wrokig with times and dates

```{r}
library(lubridate)
```

```{r}
today()
```
```{r}
today() - vax$as_of_date[1]
```

```{r}
today() -ymd("1992-8-10")
```

```{r}
today() -ymd("2021-6-28")
```

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

I will conver the entire "as_of_date" column to be in lubridate format

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```


```{r}
today() - vax$as_of_date[nrow(vax)]
```
A: 1 day!

>Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
unique <- length(unique(vax$as_of_date))
unique
```
```{r}
library(dplyr)
n_distinct(vax$as_of_date)
```
A: There are 114 unique dates in the dataset


