---
title: "Halloween Candy!"
author: "Daira"
format: pdf
---

```{r}
candy.data <- read.csv("candy-data.csv", row.names=1)
head(candy.data)

```
# what is in the dataset?

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy.data)
```
A: There are 85 candy types in this dataset

> Q2. How many fruity candy types are in the dataset?

```{r}
fruitycandy <- sum(candy.data$fruity)
fruitycandy

```
A: There are 38 fruity candies

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
fave <- candy.data["Nerds",]$winpercent
fave
```
A: My favorite candy is Nerds and it has a 55.4% winpercent value.

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
kitkat <- candy.data["Kit Kat",]$winpercent
kitkat
```
A: The winpercent value for Kit Kats is 76.8%

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
tootsie <- candy.data["Tootsie Roll Snack Bars",]$winpercent
tootsie
```
A: The win percent value for tootsie rolls is 49.7%

## skimr
```{r}
library("skimr")
skim(candy.data)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

A: Based on the skim, the win percent values seem on a different scale than the others (ie the mean, the sd..ete are larger than in a range from 0.00 to 1.00).

>Q7. What do you think a zero and one represent for the candy$chocolate column?

A: binary, for Boolean values so TRUE and FALSE, if its chocolate or not.


> Q8. Plot a histogram of winpercent values

```{r}
hist(candy.data$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

A: Not it is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
center <- median(candy.data$winpercent)
center
```
A: It appears Below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
win.choc <- candy.data[as.logical(candy.data$chocolate), "winpercent"]
win.choc
```

```{r}
win.fruit <- candy.data[as.logical(candy.data$fruity), "winpercent"]
win.fruit
```
now compare averages
```{r}
mean(win.choc)
mean(win.fruit)
```

A: On average, chocolate candy is higher ranked than fruity.

> Q12. Is this difference statistically significant?
do a sample t-test

```{r}
t.test(win.choc, win.fruit)
```

A: Yes, based on our t-test it looks like there is a significant difference between the average means (small P-value) 

```{r}
sort(c(5,4,1,2))
order(c(5,4,1,2))

```

# Overall Candy Ranking

> Q13. What are the five least liked candy types in this set?

```{r}
order.ind <- order(candy.data$winpercent)
head(candy.data[order.ind,], n=5)

```
A: The five least liked are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
order.new <-order.ind <- order(candy.data$winpercent, decreasing= TRUE)
head(candy.data[order.new,], n=5)
```
A: The top five are: ReesePeanut Butter Cup, Reese Miniatures, Twix, Kit Kat, and Snickers.

##now use ggplot2

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy.data) + 
  aes(x = winpercent, rownames(candy.data)) +
  geom_col() +
  theme_bw ()
```

```{r}
ggplot(candy.data) + 
  aes(winpercent, reorder(rownames(candy.data),winpercent)) +
  geom_col() +
  theme_bw ()
```
#can add some color

```{r}
my_cols=rep("black", nrow(candy.data))
my_cols[as.logical(candy.data$chocolate)] = "chocolate"
my_cols[as.logical(candy.data$bar)] = "brown"
my_cols[as.logical(candy.data$fruity)] = "pink"
```

```{r}
ggplot(candy.data) + 
  aes(winpercent, reorder(rownames(candy.data),winpercent)) +
  geom_col(fill=my_cols) 
```
very colorful
now we can answer some questions

> Q17. What is the worst ranked chocolate candy?

A: The worst ranked chocolate candy are Sixlets

> Q18. What is the best ranked fruity candy?

A: The best ranked fruity candy are Starbursts

## Price Percent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy.data) +
  aes(winpercent, pricepercent, label=rownames(candy.data)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

A: It appears that Reese's Miniatures are the highest rank for winpercent(80) and lower for $ 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy.data$pricepercent, decreasing = TRUE)
head( candy.data[ord,c(11,12)], n=5 )
```
A: The most expensive and worst is the Nik L Nip!

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

A: OPTIONAL 

Make a Lollipop Chart of Price percent

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy.data) +
  aes(pricepercent, reorder(rownames(candy.data), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy.data), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# Exploring Correlation

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy.data)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

A: The two variables most anti-correlated, chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

A: The most positively correlated are chocolate and win percent (and if we weren't looking at percents and other qualities of candy then it is most positively correlated with bar which makes sense because chocolate bars are delicious)

# now time for PCA

```{r}

pca <- prcomp(candy.data, scale=TRUE)
summary(pca)
```


```{r}

plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```
```{r}
ggplot(candy.data) +
aes(pca$x[,1], pca$x[,2], label=rownames(candy.data)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

```{r}
my_data <- cbind(candy.data, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r} 

#| eval: false
#| echo: false

library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (pink), other (black)",
       caption="Data from 538")
```

```{r}
#| eval: false
library(plotly)
ggplotly(p)

```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

A: The variables that are picked up by PC1 in the positive are fruity, hard, and pluribus (comes in multiple). this makes sense as fruity drives the difference in correlation of the data and msot fuirty candies are hard and come in bags of many whereas chocolate are usually in bars and in one.

```{r}
barplot(pca$rotation[,2], las=2, ylab="PC2 Contribution")
```

